### 2022-09-12 AISTATS2023/paper.tex
updated Intro

% If you use natbib package, activate the following three lines:
\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}
\bibliographystyle{plainnat}
\usepackage{amsmath}
\usepackage{amssymb}

@@

Recently, ML has flourished in critical domains, such as healthcare
and finance. In these areas, we need ML models that predict accurately
but also with the ability to explain their predictions. Therefore,
Explainable AI (XAI) is a rapidly growing field due to the interest in
interpreting black box machine learning (ML) models. XAI literature
distinguishes between local and global interpretation
methods~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the numerous local explanations into a single interpretable outcome
(number or plot). For example, if a user wants to know which features
are significant (feature importance) or whether a particular feature
has a positive or negative effect on the output (feature effect), they
should opt for a global explainability technique. Aggregating the
individual explanations for producing a global one comes at a cost. In
cases where feature interactions are strong, the global explanation
may obfuscate heterogeneous effects~\citep{Herbinger2022repid} that
exist under the hood, a phenomenon called aggregation
bias~\citep{mehrabi2021survey}.

Feature effect forms a fundamental category of global explainability
methods, isolating a single feature's average impact on the
output. Feature effect methods suffer from aggregation bias because
the rationale behind the average effect might be unclear. For example,
a feature with zero average effect may indicate that the feature has
no effect on the output or, contrarily, it has a highly positive
effect in some cases and a highly negative one in others.

There are two widely-used feature effect methods; Partial Dependence
Plots (PDPlots)\citep{friedman2001greedy} and Aggregated Local Effects
(ALE)\citep{apley2020visualizing}. PDPlots have been criticized for
producing erroneous feature effect plots when the input features are
correlated due to marginalizing over out-of-distribution synthetic
instances. Therefore, ALE has been established as the state-of-the-art
feature effect method since it can isolate feature effects in
situations where input features are highly correlated.

However, ALE faces two crucial drawbacks. First, it does not provide a
way to inform the user about potential heterogeneous effects that are
hidden behind the average effect. In contrast, in the case of PDPlots,
the heterogeneous effects can be spotted by exploring the Individual
Conditional Expectations (ICE)\citep{goldstein2015peeking}. Second,
ALE requires an additional step, where the axis of the feature of
interest is split in \(K\) fixed-size non-overlapping intervals, where
\(K\) is a hyperparameter provided by the user. This splitting is done
blindly, which can lead to inconsistent explanations.

In this paper, we extend ALE with a probabilistic component for
measuring the uncertainty of the global explanation. The uncertainty
of the global explanation expresses how certain we are that the global
(expected) explanation is valid if applied to an instance drawn at
random and informs the user about the level of heterogeneous effects
hidden behind the expected explanation. Our method completes ALE, as
ICE plots complement PDPlots, for revealing the heterogeneous effects.

Our method also automates the step of axis splitting into
non-overlapping intervals. We, firstly, transform the bin splitting
step into an unsupervised clustering problem and, second, find the
optimal bin splitting for a robust estimation of (a) the global
(expected) effect and (b) the uncertainty of the explanation from the
limited samples of the training set. We formally prove that the
objective of the clustering problem has as lower-bound the aggregated
uncertainty of the global explanation. Our method works out of the box
without requiring any input from the user.

@@

\begin{itemize}
\item We introduce Uncertainty DALE (UDALE), an extension of DALE that
  quantifies the uncertainty of the global explanation, i.e.~the level
  of heterogeneous effects hidden behind the global explanation.
\item
  We provide an algorithm that automatically computes the optimal bin
  splitting for robustly estimating the explanatory quantities, i.e.,
  the global effect and the uncertainty.
\item We formally prove that our method finds the optimal grouping of
  samples, minimizing the added uncertainty over the unavoidable
  heterogeneity that is the lower-bound of the objective.
\item We provide empirical evaluation of the method in artificial and
  real datasets.
\end{itemize}

@@

\paragraph{Feature Effect Methods.} PDPlots formulate the feature
effect of the \(s\)-th attribute as an expectation over the marginal
distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPlots formulate it
as an expectation over the conditional \(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as an
integration of local effects. The local effects are measured as the
expected change on the output
\( \frac{\partial f(x_s, \mathbf{X}_c)}{\partial x_s} \) over the
conditional distribution \( \Xcb|X_s\). The formula that
defines ALE is presented below:

\begin{equation}
  \label{eq:ALE_accumulated_mean}
  f^{\mathtt{ALE}}(x_s) = c + \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
\end{equation}
%
The constant \(c\) is used for centering the ALE plot. PDPlots
integrate over unrealistic instances due to the use of the marginal
distribution \( p(\mathcal{X}_1) \). Therefore, they incorrectly
result in a quadratic effect in the region \(x_1 \in [0, 1]\). MPlots
resolve this issue using the conditional distribution
\( \mathcal{X}_2|\mathcal{X}_1 \) but suffer from computing combined
effects. ALE plots resolve, both they have two drawbacks TODO add more
info.

>>>>

% If you use natbib package, activate the following three lines:
\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}
\bibliographystyle{plainnat}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}

@@

Recently, ML has flourished in critical domains, such as healthcare
and finance. In these areas, we need ML models that not only predict
accurately but also explain their predictions. Therefore, Explainable
AI (XAI), the field that provides interpretations for the prediction
mechanism of complex black-box models, has increased interest. XAI
literature distinguishes between local and global interpretation
methods~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the numerous local explanations into a single interpretable outcome,
usually a number or a plot. If a user wants to know which features are
significant (feature importance) or whether a particular feature has a
positive or negative effect on the output (feature effect), they
should opt for a global explainability technique. Aggregating the
individual explanations for producing a concise global one comes at a
cost. Under strong feature interactions, the global explanation may
obfuscate heterogeneous effects~\citep{Herbinger2022repid} that exist
under the hood; a phenomenon called aggregation
bias~\citep{mehrabi2021survey}.

Feature effect\citep{Gromping2020MAEP} forms a fundamental category of
global explainability methods, isolating a single feature's average
impact on the output. Feature effect methods suffer from aggregation
bias because the rationale behind the average effect might be
unclear. For example, a feature with zero average effect may indicate
that the feature has no effect on the output or, contrarily, it has a
highly positive effect in some cases and a highly negative in
others. There are three widely-used feature effect methods; Partial
Dependence Plots (PDP)\citep{friedman2001greedy}, Marignal Plots
(MP)\citep{apley2020visualizing} and Aggregated Local Effects
(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
computing erroneous effects when the input features are (highly)
correlated, which is the case in most ML setups. Therefore, ALE has
been established as the state-of-the-art feature effect method.

However, ALE faces two crucial drawbacks. First, it does not provide a
way to inform the user about potential heterogeneous effects that are
hidden behind the average effect. In contrast, in the case of PDP, the
heterogeneous effects can be spotted by exploring the Individual
Conditional Expectations (ICE)\citep{goldstein2015peeking}. Second,
ALE requires an additional step, where the axis of the feature of
interest is split in \(K\) fixed-size non-overlapping intervals, where
\(K\) is a hyperparameter provided by the user. So far, this splitting
is done in a blind way, i.e., the user does not have an indication
about the parameter K, which often leads to inconsistent explanations.

In this paper, we extend ALE with a probabilistic component for
measuring the uncertainty of the global explanation. The uncertainty
of the global explanation expresses how certain we are that the global
(expected) explanation is valid if applied to an instance drawn at
random and informs the user about the level of heterogeneous effects
behind the expected explanation. The probabilistic extension completes
ALE, as ICE plots complement PDP, for revealing the heterogeneous
effects. Our method also transforms the axis-splitting step into an
unsupervised clustering problem, i.e., we search for the optimal
splitting given the instances of the training set. We, therefore,
relieve the user from the obligation of defining a parameter \(K\) by
providing an automated method for finding the optimal bin splitting to
robustly estimate (a) the global (expected) effect and (b) the
uncertainty of the explanation, from the limited samples of the
training set.

@@

\begin{itemize}
\item We introduce Uncertainty DALE (UDALE), an extension of DALE that
  quantifies the uncertainty of the global explanation, i.e.~the level
  of heterogeneous effects hidden behind the global explanation.
\item We provide an algorithm that automatically computes the optimal
  axis-splitting for robustly estimating the the global effect and the
  uncertainty.
\item We formally prove that our method finds the optimal grouping of
  samples, minimizing the added uncertainty over the unavoidable
  heterogeneity that is the lower-bound of the objective.
\item We provide empirical evaluation of the method in artificial and
  real datasets.
\end{itemize}

@@

\paragraph{Feature Effect Methods.} PDP formulates the feature effect
of the \(s\)-th attribute as an expectation over the marginal
distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPs formulate it as
an expectation over the conditional \(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as the
accumulation of the averaged local effects, which are defined as the
change on the output, i.e.
\( \frac{\partial f(x_s, \mathbf{X}_c)}{\partial x_s} \):

\begin{equation}
  \label{eq:ALE_accumulated_mean}
  f^{\mathtt{ALE}}(x_s) = c + \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
\end{equation}
%
The constant \(c\) is used for centering the ALE plot. PDP integrates
over unrealistic instances due to the use of the marginal distribution
\( \mathbf{X}_c\). MPs resolve this issue using the conditional
distribution \( \math{C}_2|\mathcal{X}_1 \) but suffer from
computing combined effects. ALE plots resolve, both they have two
drawbacks TODO add more info.
